Pituitary prolactin drives transcriptional changes in the reproductive axis of parenting pigeons

### 2020-03-07 analysis/figure-3.Rmd
better fig 1-4
mew figs

a <- plotcolorfulpcs(charpca,charpca$treatment, allcolors) + labs(subtitle = " ") +
  theme(legend.position = "none", 
        legend.direction = "horizontal", 
        legend.key.size = unit(0.5, 'lines')) + 
  guides(color = FALSE) +
  labs(subtitle = "pituitary")

b <- plotfriz(charfviz) + labs(subtitle = "pituitary") +
  #ylim(-250000,500000) + xlim(-250000,500000) + 
  theme(axis.text = element_blank())

c <- plotprolactin(PRLpit, PRLpit$counts, "PRL", "pituitary") + 
  theme(legend.position = "none", 
         axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(face = "italic"))

@@

bcde <- plot_grid(a,b,c,d, labels = c("b", "c", "d", "e"), ncol = 2, label_size = 8)

@@

dataplots <- plot_grid(bcde, hypothesisbars)

fig3 <- plot_grid(expdesign, dataplots, rel_heights = c(0.25,1), nrow = 2, labels = c(" ", "e"), label_size = 8)
fig3

>>>>

b <- plotcolorfulpcs(charpca,charpca$treatment, allcolors) + labs(subtitle = " ") +
  theme(legend.position = "none", 
        legend.direction = "horizontal", 
        legend.key.size = unit(0.5, 'lines')) + 
  guides(color = FALSE) +
  labs(subtitle = "pituitary")

c <- plotprolactin(PRLpit, PRLpit$counts, "PRL", "pituitary") + 
  theme(legend.position = "none", 
         axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(face = "italic"))

@@

bcd <- plot_grid(b,c,d, labels = c("b", "c", "d"), ncol = 1, label_size = 8, rel_heights = c(1,1,1.3))

@@

dataplots <- plot_grid(bcd, hypothesisbars)

fig3 <- plot_grid(expdesign, dataplots, rel_heights = c(0.25,1), nrow = 2, labels = c("a"), label_size = 8)
fig3

### 2020-03-07 analysis/figure-4.Rmd
mew figs

---
title: "fig4"
output: md_document
---

# Figure 4

```{r setup, include=TRUE}
library(tidyverse)
library(cowplot)

source("../R/themes.R")

knitr::opts_chunk$set(fig.path = '../figures/',message=F, warning=FALSE)
```

```{r data}
geneids <- read_csv("../metadata/00_geneinfo.csv")

vsd_path <- "../results/DEseq2/"   # path to the data
vsd_files <- dir(vsd_path, pattern = "*vsd.csv") # get file names
vsd_pathfiles <- paste0(vsd_path, vsd_files)
vsd_files

allvsd <- vsd_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name")  %>% 
  dplyr::rename("gene" = "X1") %>% 
  pivot_longer(cols = L.G118_female_gonad_control:y98.o50.x_male_pituitary_inc.d3, 
               names_to = "samples", values_to = "counts")

hypgenes <- c("POMC", "DRD1")
pitgenes <- c("PRL", "BRCA1")
gongenes <- c("AVPR1A", "PRLR")

candidategenes <- c("PRL", "PRLR" ,  "AVPR1B" , "MYC" ,  
                "MAP2KA", "MSH" ,   "PALB2" , "BRCA1", 
                "OXT", "ESR1", "ESR2", "CISH", 
                "POU2F1", "RAP1GAP", "CEBPB","ARHGAP32"
                )

returnvsds <- function(whichgenes){

  df  <- allvsd %>%
    filter(gene %in% whichgenes) %>%
    dplyr::mutate(sextissue = sapply(strsplit(file_name, '_vsd.csv'), "[", 1)) %>%
    dplyr::mutate(sextissue = sapply(strsplit(sextissue, '../results/DEseq2/'), "[", 2)) %>%
    dplyr::mutate(sex = sapply(strsplit(sextissue, '\\_'), "[", 1),
                tissue = sapply(strsplit(sextissue, '\\_'), "[", 2),
                treatment = sapply(strsplit(samples, '\\_'), "[", 4)) %>%
    dplyr::mutate(treatment = sapply(strsplit(treatment, '.NYNO'), "[", 1)) %>%
    dplyr::select(sex, tissue, treatment, gene, samples, counts)

  df$treatment <- factor(df$treatment, levels = alllevels)
  print(head(df))
  return(df)
}

candidatevsd <- returnvsds(candidategenes)

@@

```{r fig4, fig.height=7, fig.width=7}
plottopgenes <- function(whichgenes, whichtissue, whichsex, mysubtitle, whichtstages){
  candidates  <- allvsd %>%
    filter(gene %in% whichgenes) %>%
    dplyr::mutate(sextissue = sapply(strsplit(file_name, '_vsd.csv'), "[", 1)) %>%
    dplyr::mutate(sextissue = sapply(strsplit(sextissue, '../results/DEseq2/'), "[", 2)) %>%
    dplyr::mutate(sex = sapply(strsplit(sextissue, '\\_'), "[", 1),
                tissue = sapply(strsplit(sextissue, '\\_'), "[", 2),
                treatment = sapply(strsplit(samples, '\\_'), "[", 4)) %>%
    dplyr::mutate(treatment = sapply(strsplit(treatment, '.NYNO'), "[", 1)) %>%
    dplyr::select(sex, tissue, treatment, gene, samples, counts) %>%
    filter(tissue == whichtissue, sex %in% whichsex)

  candidates$treatment <- factor(candidates$treatment, levels = alllevels)

  p <- candidates %>%
    filter(treatment %in% whichtstages) %>%
    ggplot(aes(x = treatment, y = counts, fill = treatment, color = sex)) +
    geom_boxplot() + facet_wrap(~gene, scales = "free_y") + 
    theme_B3() +
    scale_fill_manual(values = allcolors) +
    scale_color_manual(values = allcolors) +
    theme(axis.text.x = element_blank(),
          legend.position = "none",
          axis.title.x = element_blank(),
          strip.text = element_text(face = "italic")) +
    labs(y = mysubtitle) 
  return(p)

}

a <- plottopgenes(hypgenes, "hypothalamus", c("female"), "hypothalamus", c("hatch", "n5")) 
b <- plottopgenes(pitgenes, "pituitary", c("female"), "pituitary", c("inc.d9", "inc.d17")) 
c <- plottopgenes(gongenes, "gonad", c("female"), "gonads", c("lay", "inc.d3")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fig4a <- plot_grid(a,b,c, ncol = 1)

d <- plottopgenes(hypgenes, "hypothalamus", c("female", "male"), NULL, charlevels) 
e <- plottopgenes(pitgenes, "pituitary", c("female", "male"), NULL, charlevels) 
f <- plottopgenes(gongenes, "gonad", c("female", "male"), NULL, charlevels) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fig4b <- plot_grid(d,e,f, ncol = 1)

fig4 <- plot_grid(fig4a, fig4b, rel_widths = c(1,2))
fig4
```

```{r savefiles}
write.csv(candidatevsd, "../../musicalgenes/data/candidatecounts.csv")
```

>>>>

---
title: "Fig5"
output: md_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
library(Rtsne)

source("../R/themes.R")
source("../R/functions.R")
source("../R/wrangledata.R")

knitr::opts_chunk$set(echo = TRUE, fig.path = '../figures/')
```

# Figure 1. Experimental design, tSNE analysis, PCA, and prolactin

```{r tSNE-df}

# prep for tsne

hyptsne <- subsetmaketsne("hypothalamus", allmaniplevels, sexlevels)
pittsne <- subsetmaketsne("pituitary", allmaniplevels, sexlevels)
gontsne <- subsetmaketsne("gonads", allmaniplevels, sexlevels)

addgroupings <- function(df){

  df <- df %>% mutate(hiloPRL = fct_collapse(treatment, 
                                  lo = c("m.inc.d8", "m.inc.d3", "m.inc.d9", "inc.d3", "inc.d9", "lay", "bldg"),
                                  hi = c("inc.d17",   "m.inc.d17", "prolong" ,  "hatch" ,  "m.n2", "extend", "n5")),
                extint = fct_collapse(treatment, 
                                  eggs = c("inc.d3", "inc.d9", "inc.d17", "prolong", "lay"),
                                  chicks = c( "m.inc.d8", "hatch", "extend", "n5"),
                                  loss = c("m.inc.d3", "m.inc.d9", "m.n2", "m.inc.d17", "bldg")))
  df$extint <- factor(df$extint, levels = levelsextint)
  return(df)
}

hyptsne <- addgroupings(hyptsne)
pittsne <- addgroupings(pittsne)
gontsne <- addgroupings(gontsne)

@@

## make figure

```{r fig4, fig.width=7, fig.height=7}

expdesign2 <- png::readPNG("../figures/images/fig_fig4a.png")
expdesign2 <- ggdraw() +  draw_image(expdesign2, scale = 1)

b <- plottsneelipsev2(hyptsne, hyptsne$treatment, allcolors) + labs(subtitle = "hypothalamus ~ parental stage", x = NULL)  + facet_wrap(~sex, scales = "free")
c <- plottsneelipsev2(pittsne, pittsne$treatment, allcolors ) + labs(subtitle = "pituitary ~ parental stage", x = NULL) + facet_wrap(~sex, scales = "free") 
d <- plottsneelipsev2(gontsne, gontsne$treatment, allcolors ) + labs(subtitle = "gonads ~ parental stage")  + facet_wrap(~sex, scales = "free")

bcd <- plot_grid(b,c,d, nrow = 3, labels = c("b"), label_size = 8, rel_heights = c(1,1,1))

e <- plottsneelipsev2(hyptsne, hyptsne$hiloPRL, allcolors ) + labs(subtitle = "hypothalamus ~ internal physiology", x = NULL) + facet_wrap(~sex, scales = "free")  
f <- plottsneelipsev2(pittsne, pittsne$hiloPRL, allcolors ) + labs(subtitle = "pituitary ~ internal physiology", x = NULL) + facet_wrap(~sex, scales = "free")   
g <- plottsneelipsev2(gontsne, gontsne$hiloPRL, allcolors ) + labs(subtitle = "gonads ~ internal physiology" ) + facet_wrap(~sex, scales = "free")

efg <- plot_grid(e,f,g, nrow = 3, labels = c("d"), label_size = 8, rel_heights = c(1,1,1))

h <- plottsneelipsev2(hyptsne, hyptsne$extint, allcolors ) + labs(subtitle = "hypothalamus ~ external enviornment", x = NULL) + facet_wrap(~sex, scales = "free")  
i <- plottsneelipsev2(pittsne, pittsne$extint, allcolors ) + labs(subtitle = "pituitary ~ external enviornment", x = NULL) + facet_wrap(~sex, scales = "free")   
j <- plottsneelipsev2(gontsne, gontsne$extint, allcolors ) + labs(subtitle = "gonads ~ external enviornment" ) + facet_wrap(~sex, scales = "free")

hij <- plot_grid(h,i,j, nrow = 3, labels = c("c"), label_size = 8, rel_heights = c(1,1,1))

bcdhijefg <- plot_grid(bcd, hij,  efg, nrow = 1)

fig5 <- plot_grid(expdesign2, bcdhijefg, nrow = 2, labels = c("a"), label_size = 8, rel_heights = c(0.25,1))
fig5

### 2020-03-07 docs/manuscript/elife_v1.Rmd
mew figs

---
title: Internal physiology and external cues drive changes in reproductive axis of parenting pigeons
author:
  - name: Rayna M Harris ^[equal contribtion] 
    affiliation: UC Davis
  - name: Suzanne H. Austin  ^[equal contribtion] 
    affiliation: UC Davis
  - name: Andrew S. Lang
    affiliation: UNH
  - name: Victoria S. Farrar
    affiliation: UC Davis
  - name: April Booth
    affiliation: UC Davis
  - name: Tanner Feustel
    affiliation: UC Davis
  - name: Matthew D. MacManes
    affiliation: UNH
  - name: Rebecca M. Calisi  ^[corresponding author] 
    affiliation: UC Davis
    email:  author7@work   
address:
  - code: UC Davis
    address: Department, Street, City, CA, Zip
  - code: UNH
    address: Department, Street, City, NH, Zip    
abstract: |
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

@@

@@

Our results suggest that tissue is the greatest source of variation in the dataset, with hypothalamic, pituitary, and gonadal samples all forming distinct clusters, with only a few pituitary and gonad samples falling outside the 95% confidence interval (Fig. \ref{figure:fig1}b). Sex-differences are most prounounced in the gonads with male and female samples form two discrete clusters (Fig. \ref{figure:fig1}c). Treatment accounts for much less of the variation than tissue and sex, and even after removing sex as a variatble, treatment effects in all three tissues are hard to visualize simultatenously (Fig. \ref{figure:fig1}d,e), so we next looked at each tissue and sex separately. In the hypothalamus, control samples form tight clusters that stands out from all other timepoints that are largely overlapping in space (Fig. \ref{figure:fig1}h). The pituitary shows the strongest treatment effect with one cluster of the “earlier” parental timepoints (nest-building, lay, incubation days 3 and 9), a second cluster of “later” parental timpoints (incubation day 17, hatch, and nestling care days 5 and 9; Fig. \ref{figure:fig1}i). Like the hypothalamus, in the gonads, the control timepoints appear to be much less variable that the parental timepoints, but there is no clear separation of gonadal samples by parental stage (Fig. \ref{figure:fig1}j). While the tSNE analysis reveals broad patterns across all levels of the experiment, additional tools are needed it hone in on the specific differences associated with parental care in each tissue.

Differential gene expression analysis can shed insight into specific differences between treatment groups and thus deepen our understand of the transitions between reproduction and parental care behaviors. We used DESEq2 to identify genes whose expression changed significantly between sequential timepoints in each tissue and sex separately (Fig. \ref{figure:fig2}a).

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig1-1}
  \caption{Broad patterns of variation in hypothalamic, pituitary, and gonadal gene expression in male and female rock doves across reproductive and parental transitions. a) Experimental design. b) t-SNE plot highlighting tissue differences in gene expression. Each dot represents 1 sample and is colored by the source tissue (hypothalamus: orange, pituitary: teal, gonads: purple). c) t-SNE plot highlighting sex differences in tissue and sex, as shown in light grey (females) and dark grey (males). t-SNE plots highlight treatment effects in females (d), males (e), the hypothalamus (h), pituitary (i), and gonads (j), with samples colored by parental stage (control: red or , nest-building: orangse, lay: yellow; inc.d3: light green, inc.d9: green, inc.d17: dark green, hatch: dark blue, n5: blue, and n9: light blue.}
  \label{figure:fig1}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig4-1}
  \caption{Another figure}
  \label{figure:fig4}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig5-1}
  \caption{Another another figure}
  \label{figure:fig5}
\end{figure}

@@

Results in context to literature.

Significance.

Limitations.

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/supplfig3}
  \caption{R shiny, image 1}
  \label{figure:supplfig3}
\end{figure}

@@

### Animal Care

@@

### Experimental design and tissue processing

@@

### RNA sequencing analysis

@@

### Data availablility

@@

>>>>

---
title: Pituitary prolactin drives transcriptional changes in the reproductive axis of parenting pigeons
author:
  - name: Rayna M Harris ^[equal contribtion] 
    affiliation: University of California, Davis
  - name: Suzanne H. Austin  ^[equal contribtion, currently at the University of Oregon] 
    affiliation: University of California, Davis
  - name: Andrew S. Lang
    affiliation: University of New Hampshire
  - name: Victoria S. Farrar
    affiliation: University of California, Davis
  - name: April Booth
    affiliation: University of California, Davis
  - name: Tanner Feustel
    affiliation: University of California, Davis
  - name: Matthew D. MacManes
    affiliation: University of New Hampshire
  - name: Rebecca M. Calisi  ^[corresponding author] 
    affiliation: University of California, Davis
    email:  rmcalisi@ucdavis.edu  
address:
  - code: University of California, Davis
    address: Department of Neurobiology, Physiology, and Behavior, Davis, CA
  - code: University of New Hampshire
    address: Department of Molecular, Cellular and Biomedical Sciences, Durham, NH 
abstract: In species that provide parental care, successful rearing of offspring involves a shift from aggressive and sexual behaviors to more caring and nurturing ones.  We hypothesize that these shifts in behavior are related to changes in gene expression that are controlled by internal mechanisms rather than by the external cues.  We found that gene expression in the hypothalamus is fairly insensitive to the natural transition between egg and nestling stages; however, in the pituitary and gonads, gene expression showed much more plasticity over the course of reproduction. We show evidence that natural transitions in reproductive and parental behavior by mediated internal clock controlling gene expression in the pitary and gonads but that the hypothalamus is very sensitive to external cues that signal a distruption in the natural parental care cycle. This research provides a deeper understanding of the interplay of genes, hormones, and parental care and has important implications for understanding molecular processes that underlie behavior transitions. 
output:
  rticles::rjournal_article
---

@@

_Characterization of parental care_

@@

Our results suggest that tissue is the greatest source of variation in the dataset, with hypothalamic, pituitary, and gonadal samples all forming distinct clusters, with only a few pituitary and gonad samples falling outside the 95% confidence interval (Fig. \ref{figure:fig1}b). Sex-differences are most prounounced in the gonads with male and female samples form two discrete clusters (Fig. \ref{figure:fig1}c). Treatment accounts for much less of the variation than tissue and sex, and even after removing sex as a variatble, treatment effects in all three tissues are hard to visualize simultatenously (Fig. \ref{figure:fig1}d,e), so we next looked at each tissue and sex separately. In the hypothalamus, control samples form tight clusters that stands out from all other timepoints that are largely overlapping in space (Fig. \ref{figure:fig1}h). The pituitary shows the strongest treatment effect with one cluster of the “earlier” parental timepoints (nest-building, lay, incubation days 3 and 9), a second cluster of “later” parental timpoints (incubation day 17, hatch, and nestling care days 5 and 9; Fig. \ref{figure:fig1}i). Like the hypothalamus, in the gonads, the control timepoints appear to be much less variable that the parental timepoints, but there is no clear separation of gonadal samples by parental stage (Fig. \ref{figure:fig1}j). While the tSNE analysis reveals broad patterns across all levels of the experiment, additional tools are needed it hone in on the specific differences associated with parental care in each tissue. Differential gene expression analysis can shed insight into specific differences between treatment groups and thus deepen our understand of the transitions between reproduction and parental care behaviors. We used DESEq2 to identify genes whose expression changed significantly between sequential timepoints in each tissue and sex separately (Fig. \ref{figure:fig2}a).

@@

_Internal physiology versus external stimuli hypothesis_

We show that samples from the pituitary on 
                     incubation day 17 cluster more closely with hatch and nestling care 
                     day 5 samples than with incubation days 3 and 9. This suggest that, 
                     in the pituitary, internal mechanisms that prepare for chick arrival 
                     have the greatest effect on gene expression rather than suggesting 
                     that gene expression responds to changes in the external enviornament.

Blah \ref{figure:fig3}, blah, \ref{figure:fig4} blah.

_Potential implications for cancer research_

These results confirmed prolactin (PRL) gene expression fluctuates throughout parental care 
in a manner that is consistent with its role in promoting lactation 
                and maintaining parental behaviors.
                Interestingly, we identified about 100 genes whose expression 
                was correlated with PRL,
                including (BRCA1, which is associated with breast cancer).
                Additionally, we identified thousands of genes whose expression 
                changed over 
                the course of parental care.
                Exploring the relationship between PRL and other differentially 
                expressed genes could provide important insights into the `symphony` of 
                gene expression 
                that regulates behavior at the organismal and cellular level.

_Software for data exploration and hypothesis testing_

@@

_Results in context to literature. _

blah blah balh

_Significance. _

blah blah balh

_Limitations._

blah blah balh

@@

_Animal Care_

@@

_Experimental design and tissue processing_

@@

_RNA sequencing analysis_

@@

_Data availablility_

@@

\newpage

\newpage

\newpage

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig3-1}
  \caption{Another figure}
  \label{figure:fig3}
\end{figure}

\newpage

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig4-1}
  \caption{Another another figure}
  \label{figure:fig4}
\end{figure}

\newpage

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/supplfig3}
  \caption{R shiny, image 1}
  \label{figure:supplfig3}
\end{figure}

\newpage

## Author affiliations
